---
title: "tberf"
author: "Maggi Brisbin"
date: "2025-12-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Set up: import packages and define color palette*
```{r import packages, warning=FALSE, message=FALSE}
library(dplyr) # dataframe manipulation
library(stringr)
library(ggplot2) # plotting package
library(lubridate)
library(tidyr)
library(scales)
library(ggOceanMaps)
library(ggspatial)
`%ni%` = Negate(`%in%`)

TBcolors <- c("#9e0142", "#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#66c2a5", "#3288bd", "#5e4fa2"  )
```


*Make Florida map with square designating the bounds of themap for the OTB sampling region*
```{r}
#define limits of map
dt <- data.frame(lon = c(-79.5, -79.5, -85, -85), lat = c(25, 30.5, 30.5, 25)) 

FL <- basemap(data = dt, bathymetry = TRUE) + 
   geom_rect(aes(xmin = -82.9, xmax = -82.3, ymin = 27.65, ymax = 28.1), color = "black", fill = NA)  + xlab("") + ylab("") + 
  ggspatial::annotation_scale(location = "bl", line_width = 1,
  height = unit(0.5, "cm"),
  pad_x = unit(1, "cm"),
  pad_y = unit(0.5, "cm"),
  text_pad = unit(0.3, "cm"),
  text_cex = 1,
  text_face = NULL,
  text_family = "",
  tick_height = 1) + 
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true",  height = unit(1.5, "cm"),
  width = unit(1.5, "cm")) 

FL
```

ggsave("FLmap.png", width = 4, height = 4)

*Read in FWC data from OTB sampling in the summer of 2025*
- this data was acquired through a request made to HABdata@myFWC.com 

```{r}
FWC <- read.csv("TBPhyto20250508-20251014.csv") #this is the raw data file provided by FWC
FWC$Station_Number <-  sapply(strsplit(FWC$Station.ID, "\\s+"), tail, 1) # formatting station numbers
FWC$Station_Number <- factor(FWC$Station_Number, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10")) # formatting station numbers
```


*Plots sample locations and label station number*

```{r}
locs <- unique(FWC %>% select(Station_Number, Latitude, Longitude)) #create smaller data frame with one row per station and including only Station Number, Lat, Long

dt <- data.frame(lon = c(-82.3, -82.3, -82.9, -82.9), lat = c(27.65, 28.1, 28.1, 27.65)) #define limits of OTB map

OTB <- basemap(data = dt, bathymetry = TRUE)   + geom_point(data = locs, aes(x = Longitude, y = Latitude, color = Station_Number))+ xlab("") + ylab("") + theme(legend.position = "none") +  scale_color_manual(values = TBcolors) + geom_text(data = locs, aes(x = Longitude, y = Latitude, label = Station_Number), nudge_x = 0.0215) + ggtitle("OTB sampling stations")

OTB
```
ggsave("OTBstations.png", width = 4, height = 4)

*Spatial view of data*
*Plot Salinity data spatially for one time point:* 
```{r}
OTBsalinity <- basemap(data = dt, bathymetry = TRUE)   + geom_point(data = (sal %>% filter(Sample.Date == "2025-05-08")), aes(x = Longitude, y = Latitude, color = Salinity), size = 3)+ xlab("") + ylab("") + theme(legend.position = "right")  + geom_text(data = locs, aes(x = Longitude, y = Latitude, label = Station_Number), nudge_x = 0.025) + scale_color_continuous(type = "viridis") + ggtitle("2025-05-08")

OTBsalinity
```
- This is better in a dynamic Shiny App, rather than plotting one date at a time
- See: https://micolab-usf.shinyapps.io/otbact/

*Time-series plotting*

*Salinity* 
```{r}
sal <- unique(FWC %>% select(Sample.Date, Sample.Depth..m., Station_Number, Latitude, Longitude, Salinity)) %>%  filter(Sample.Depth..m. == 0.5) #subset FWC datafrome to only include Salinity at the near surface

sal$Sample.Date <- dmy(sal$Sample.Date) # format date column

salP<- sal %>%  ggplot(aes(x= Sample.Date, y = Salinity, color = Station_Number)) + geom_point() + geom_line() + scale_color_manual(values = TBcolors) + theme_test()

salP
```



*Temperature*
```{r}
temp <- unique(FWC %>% select(Sample.Date, Sample.Depth..m., Station_Number, Latitude, Longitude, Temp..ᵒC.)) %>%  filter(Sample.Depth..m. == 0.5)

temp$Sample.Date <- dmy(temp$Sample.Date)

tempP <- temp %>%  ggplot(aes(x= Sample.Date, y = Temp..ᵒC., color = Station_Number)) + geom_point() + geom_line() + scale_color_manual(values = TBcolors) + theme_test()

tempP
```
ggsave("salinity.pdf", width = 6, height = 3)

*Chlorophyll-a fluorescence* 

```{r}
chl <- unique(FWC %>% select(Sample.Date, Sample.Depth..m., Station_Number, Latitude, Longitude, Chlorophyll.fluorescence..µg.L.)) %>%  filter(Sample.Depth..m. == 0.5)

chl$Sample.Date <- dmy(chl$Sample.Date)

chlP<- chl %>%  ggplot(aes(x= Sample.Date, y = Chlorophyll.fluorescence..µg.L., color = Station_Number)) + geom_point() + geom_line() + scale_color_manual(values = TBcolors) + theme_test()

chlP
```

*Bacterial Cell Counts*

```{r warning=FALSE}
bact = data.frame(matrix(
  vector(), 0, 13, dimnames=list(c(), c("Sample", "Station" , "Rep" , "Gate", "X.Parameter", "Y.Parameter",  "Volume", "Concentration", "X.Mean", "Y.Mean", "X.SD", "Y.SD", "Date"))),
                stringsAsFactors=F)

for (i in system("ls OTB_Sampling_2025/", intern = TRUE)) {
  df<- read.csv(paste0("OTB_Sampling_2025/",i))
  date <- str_split_i(i, "_", 3)
  date2<- str_extract(date, '.*(?=\\.csv$)')
  df<- df %>% separate(Sample, into = c("Station", "var1", "var2"), sep = "_", remove = FALSE) %>% mutate(Rep = case_when(
  var1 == "" ~ var2,
  var1 != "" ~ var1
)) %>% 
  #filter(Gate =="R3") %>% 
  select(Sample, Station, Rep, Gate, X.Parameter, Y.Parameter, Volume, Concentration, X.Mean, Y.Mean, X.SD, Y.SD) %>% 
  filter(Sample != "Blank") 
  df$Date <- date2
  bact <- rbind(bact, df)
}

bact$Date <- ymd(bact$Date)


bact<- bact %>% mutate(Station_Number = case_when(
  Station == "St1" ~ "01", 
  Station == "St2" ~ "02",
  Station == "St3" ~ "03",
  Station == "St4" ~ "04",
  Station == "St5" ~ "05",
  Station == "St6" ~ "06",
  Station == "St7" ~ "07",
  Station == "St8" ~ "08",
  Station == "St9" ~ "09",
  Station == "St10" ~ "10"
))

bact$Station_Number <- factor(bact$Station_Number, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10"))
```



filter(Gate == "R3") 
```{r}
bact4plot <- bact %>% filter(Gate == "R3") %>% filter(X.Parameter =="FSC - FSC-H")

bact4plot %>% ggplot(aes(x=Date, y = Concentration, color = Station_Number)) + geom_point() + theme_test() + scale_color_manual(values = TBcolors)
```



```{r}
bact4plot %>% group_by(Station_Number, Date,) %>% dplyr::summarize(mean_conc = mean(Concentration), sd = sd(Concentration)) %>% ggplot(aes(x=Date, y = mean_conc, color = Station_Number)) + geom_point() + theme_test() + scale_color_manual(values = TBcolors) + geom_errorbar(aes(ymin=mean_conc-sd, ymax=mean_conc+sd), width=.001) 
```


```{r}

bactP<- bact4plot %>% group_by(Station_Number, Date,) %>% dplyr::summarize(mean_conc = mean(Concentration), sd = sd(Concentration)) %>% ggplot(aes(x=Date, y = mean_conc, color = Station_Number)) + geom_point() + theme_test() + scale_color_manual(values = TBcolors) + geom_errorbar(aes(ymin=mean_conc-sd, ymax=mean_conc+sd), width=.001) + geom_line()

bactP
```

ggsave("BacteriaCounts.pdf", width = 6, height = 3)


Combine Data and Correlate
```{r}
field <- unique(FWC %>% filter(Sample.Depth..m. == 0.5) %>% select(Sample.Date, Station.ID, Latitude, Longitude, Chlorophyll.fluorescence..µg.L., Temp..ᵒC., Salinity)) 

field$Station_Number <-  sapply(strsplit(locs$Station.ID, "\\s+"), tail, 1)

field <- field %>% select(Station_Number, Sample.Date, Chlorophyll.fluorescence..µg.L., Temp..ᵒC., Salinity, Latitude, Longitude)
names(field) <- c("Station_Number", "Date", "Chla", "Temp", "Salinity", "Latitude", "Longitude" )

field$Date <- dmy(field$Date )

bact4merge<- bact4plot %>% group_by(Station_Number, Date) %>% dplyr::summarize(mean_conc = mean(Concentration))

field$Station_Number <- factor(field$Station_Number, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10"))

joined <- full_join(field, bact4merge)

#write.csv(joined, "CleanedData.csv", quote = FALSE, row.names = FALSE)
```



```{r}
#test
joined%>% ggplot(aes(x=Date, y = Chla, color = Station_Number)) + geom_point() + theme_test() + scale_color_manual(values = TBcolors)  + geom_line()
```


```{r}
library(patchwork)

tempP / salP / bactP + plot_layout(guides = "collect")
```

ggsave("temp_sal_bact.pdf",width = 6, height = 8)


```{r}
library(magrittr)
library(liver)

joined$Temp_z <- transform(joined$Temp, method = "zscore")
joined$Chla_z  <- transform(joined$Chla, method = "zscore")
joined$Salinity_z  <- transform(joined$Salinity, method = "zscore")
joined$Bact_z  <- transform(joined$mean_conc, method = "zscore")

Zs<- joined %>% select(Temp_z, Chla_z, Salinity_z, Bact_z)
```


```{r}
library(corrplot)

testRes = cor.mtest( Zs, conf.level = 0.95)


M = cor(Zs, use = "na.or.complete",
    method = "spearman")

corrplot(M, p.mat = testRes$p, diag = FALSE, type = 'full', 
         sig.level = c( 0.01), pch.cex = 1.3,
         insig = 'label_sig', pch.col = 'black', tl.col = "black", title = "")

```

```{r}
M
```


----- 
Make Matrix groups 

```{r}
ext_dates <- joined %>% filter(Date %in% c("2025-05-08", "2025-06-19", "2025-07-01", "2025-07-31", "2025-08-29", "2025-09-25", "2025-10-14"))


#70 samples. 70 - 6 = 64 
# 64/8 = 8
# so 9 matrix groups total 

matrix<- ext_dates %>% select(Station_Number, Date, Salinity, Chla, mean_conc, Temp) %>%  arrange(Salinity)

matrix$order <- c(1:70)

write.csv(matrix, "matrixorder.csv", quote = FALSE, row.names = FALSE)
```


*Microscopy-based taxonomy and cell counts*
```{r}
taxa <- FWC%>%  filter(Sample.Depth..m. == 0.5) %>% select(Sample.Date, Sample.Depth..m., Station_Number, Latitude, Longitude, Taxa, Abundance..cells.L..) %>% filter(Abundance..cells.L.. != 0)

taxa$Sample.Date <- factor(taxa$Sample.Date, levels = c("08-May-25", "23-May-25" ,"11-Jun-25", "19-Jun-25", "01-Jul-25", "18-Jul-25", "31-Jul-25", "13-Aug-25", "29-Aug-25","11-Sep-25" ,"25-Sep-25" ,"14-Oct-25")) 
```

```{r}
taxa %>% ggplot( aes(x = Station_Number, y = Abundance..cells.L.., fill = Taxa)) +
  geom_bar(stat = "identity") + facet_grid(Sample.Date~.)
```

```{r}
taxa %>% ggplot( aes(x = Station_Number, y = Abundance..cells.L.., fill = Taxa)) +
  geom_bar(stat = "identity") + facet_grid(Sample.Date~., scales = "free_y")
```


Only to Genus level: 

```{r}
color_palette_25 <- c(
  "#E6194B", "#3CB44B", "#4363D8", "#FFE119", "#F58231",
  "#911EB4", "#46F0F0", "#F032E6", "#BCF60C", "#FABEBE",
  "#008080", "#E6BEFF", "#9A6324", "#FFFAC8", "#800000",
  "#AAFFC3", "#808000", "#FFD8B1", "#000075", "#A9A9A9",
  "lightblue", "#FF6B6B", "green", "#DDA0DD", "#2E8B57", "black")
```


```{r}
taxa$taxa_1stword <- sapply(strsplit(taxa$Taxa, "\\s+"), head, 1) 

taxa %>% ggplot( aes(x = Station_Number, y = Abundance..cells.L.., fill = taxa_1stword)) +
  geom_bar(stat = "identity") + facet_grid(Sample.Date~., scales = "free_y") + scale_fill_manual(values = color_palette_25) + theme_classic()

```
Relative Abundance: 

```{r}
taxa_RA<- taxa %>% group_by(Station_Number, Sample.Date) %>% mutate(Abundance_total = sum(Abundance..cells.L..)) %>% mutate(RA = (Abundance..cells.L../Abundance_total)*100)
```

```{r}
taxa_RA %>% ggplot( aes(x = Station_Number, y = RA, fill = taxa_1stword)) +
  geom_bar(stat = "identity") + facet_grid(Sample.Date~., scales = "free_y") + scale_fill_manual(values = color_palette_25) + theme_classic()
```


